Choose a License

### 2024-11-26 choose_license.qmd
Improve wording
Improve on license choosing

Every expression of an idea, such as a literary or an artistic works, is automatically covered by copyright. This means that nobody else than the copyright holder is allowed to copy, modify, or share it. Copyright licenses can grant others some of the necessary rights, but mostly don't cover publicity, privacy, moral, patent, or trademark rights.

@@

When you have your article published under an open access license, some publishers still demand an exclusive publishing and distribution license or a copyright assignment. This would give them more rights than the readers of the article have through the chosen open access license [@Rumsey2022] and exceeds exceeds by far what is necessary to make publication possible [@Suber2022]. Consequently, authors should oppose this practice and grant publishers the same rights that every other reader of the article has. If your chosen publisher insists on an exclusive license, you may at least retain the copyright for your figures -- follow the guide "Retaining copyright for figures in academic publications to allow easy citation and reuse" by @Elson2016 to learn how to do that.

@@

If you create a new work and no strong community norms suggest a particular license, you need to choose the license yourself. Which license to choose depends on the type of work you create. Software licenses, for example, may consider that the source code is the preferred form for making modifications, while licenses for data can differentiate between the database and any works produced from it. We have created a flowchart that covers the most likely types of works you will create as a researcher: software, writing (i.e., text), images, audio, video, and data (see @fig-flowchart-simple). This flowchart always recommends the most permissive license possible to maximize reuse -- below we provide two additional flowcharts that allow for more choices. Click on the name of a license to learn more about it.

@@

Note, that for data you have, at least in principle, the option to apply separate licenses to the individual entries and the collective database. For example, if you were to create a database of artworks by others, those artworks would be licensed individually as chosen by the artists, but the license for the database as a whole could be chosen by you. The latter includes the structure of the database (e.g., the selection of entries and field names) and any _sui generis_ database rights. However, if the content was created by you, we recommend you to choose the same license for both content and database. Metadata, in particular, should always be licensed under [CC0\ 1.0](https://creativecommons.org/publicdomain/zero/1.0/).

@@

- __fonts:__ Copyleft licenses applied to fonts can a special case: If a font is put under the license [CC\ BY-SA\ 4.0](https://creativecommons.org/licenses/by-sa/4.0/), any documents containing texts using that font will probably be derivative works and have to be put under the same license if shared. If the intent is that only derived fonts, if published, have to be put under the same license, the [SIL Open Font License\ 1.1](https://openfontlicense.org/) is an appropriate choice. Note, however, that it doesn't require attribution for usage. If no copyleft mechanism is intended, [CC0\ 1.0](https://creativecommons.org/publicdomain/zero/1.0/) also works for fonts.
- __templates and LaTeX packages:__ If a template or, for this purpose, a LaTeX package, is licensed under a copyleft software license such as the [AGPLv3](https://choosealicense.com/licenses/gpl-3.0/), every work that is a derivative of the template or that uses the LaTeX package has to put under the same license, if shared [@Koppor2016]. And if a document contained source code covered by the GPLv3, the same license would also apply to the document [@MadHatter2022].
- __database content:__ If the work produced from a database (the "output") is a derivative of the content in the database, the output is subject to the restrictions laid out in the license. For example, if geospatial data were to be licensed under [CC\ BY\ 4.0](https://creativecommons.org/licenses/by/4.0/), all maps produced from the data would likely need to fulfill this license's obligation for unrestricted access if shared [see @Poole2017]. Similarly, following an example from @Matt2009 if one were to choose the copyleft license [CC\ BY-SA\ 4.0](https://creativecommons.org/licenses/by-sa/4.0/) for this purpose, any map that is a derivative of the data would also need to licensed under CC\ BY-SA\ 4.0 (or a compatible license) if shared. If the intention is to only have [derivative databases](https://osmfoundation.org/wiki/Licence/Community_Guidelines/Produced_Work_-_Guideline) under the same license, one might want to choose the [ODbL\ 1.0](https://opendatacommons.org/licenses/odbl/summary/) for the database, as it was specifically designed not to apply to works produced from the data in the database. Otherwise [CC0\ 1.0](https://creativecommons.org/publicdomain/zero/1.0/) is an excellent choice for data.
- __works in the public domain__: If a work is already in the public domain, it should be marked using the [PDM 1.0](https://creativecommons.org/publicdomain/mark/1.0/), rather than applying a waiver such as the [CC0\ 1.0](https://creativecommons.org/publicdomain/zero/1.0/) (or even another license).
:::

@@

The consequence of applying a permissive license to your work is that others may distribute their modifications to it under stricter terms, possibly without a free/open license. You may prefer that everybody who receives somebody else's adaptations of your work enjoys the same rights that you granted to them. This can be achieved by _copyleft_ licenses.^[So called because copyleft licenses make content permanently free, thus turning copyright around: © → 🄯. In the realm of Creative Commons licenses, copyleft is also called _share alike_.] For software, they come in two flavors:

@@

The answer is that you need to indicate the license on a per-file or per-folder basis (rather than choosing one for the whole project). The easiest approach is to make note of every foreign work included in your project and record its license. Then, multi-license all the remaining files, which are yours, under a code and a non-code license in parallel as explained in @tip-license-r.

>>>>

Every expression of an idea, such as a literary or an artistic works, is automatically covered by copyright. Put simply, this means that nobody else than the copyright holder is allowed to copy, modify, or share it. Copyright licenses can grant others some of the necessary rights, but mostly don't cover publicity, privacy, moral, patent, or trademark rights. Certain exceptions and limitations to copyright (such as fair use or the right to quote) may obviate the need for a license.

@@

When you have your article published under a Creative Commons license, some publishers still demand an exclusive publishing and distribution license or a copyright assignment. This would give them more rights than the readers of the article have through the chosen Creative Commons license [@Rumsey2022] and exceeds by far what is necessary to make publication possible [@Suber2022]. Consequently, authors should oppose this practice and grant publishers the same rights that every other reader of the article has. If your chosen publisher insists on an exclusive license, you may at least retain the copyright for your figures -- follow the guide "Retaining copyright for figures in academic publications to allow easy citation and reuse" by @Elson2016 to learn how to do that.

@@

If you create a new work and no strong community norms suggest a particular license, you need to choose the license yourself. Which license to choose depends on the type of work you create. Software licenses, for example, may consider that the source code is the preferred form for making modifications, while licenses for data can differentiate between the database and any works produced from it. We have created a flowchart that covers the most likely types of works you will create as a researcher: software, writing (i.e., text), images, audio, video, and data (see @fig-flowchart-simple). This flowchart always recommends the most permissive license possible to maximize reuse -- though we provide two additional flowcharts below that allow for more choices. Click on the name of a license to learn more about it.

@@

Note, that for data you have, at least in principle, the option to apply different licenses to the individual entries and the collective database. For example, if you were to create a database of artworks by others, those artworks would be licensed individually as chosen by the artists, but the license for the database as a whole could be chosen by you. The latter includes the structure of the database (e.g., the selection of entries and field names) and any _sui generis_ database rights. However, if the content was created by you, we recommend you to choose the same license for both content and database. Metadata, in particular, should always be licensed under [CC0\ 1.0](https://creativecommons.org/publicdomain/zero/1.0/).

@@

- __fonts:__ Copyleft licenses applied to fonts can be a special case: If a font is put under the license [CC\ BY-SA\ 4.0](https://creativecommons.org/licenses/by-sa/4.0/), any documents containing texts using that font will probably be derivative works and have to be put under the same license if shared. If the intent is that only derived fonts, if published, have to be put under the same license, the [SIL Open Font License\ 1.1](https://openfontlicense.org/) is an appropriate choice. Note, however, that it doesn't require attribution for usage. If no copyleft mechanism is intended, [CC0\ 1.0](https://creativecommons.org/publicdomain/zero/1.0/) also works for fonts.
- __templates and LaTeX packages:__ If a template or, for this purpose, a LaTeX package, is licensed under a copyleft software license such as the [AGPLv3](https://choosealicense.com/licenses/gpl-3.0/), every work that is a derivative of the template or that uses the LaTeX package has to put under the same license, if shared [@Koppor2016]. And if a document contained source code covered by the GPLv3, the same license would also apply to the document [@MadHatter2022].
- __database content:__ If the work produced from a database (the "output") is a derivative of the content in the database, the output is subject to the restrictions laid out in the license. For example, if geospatial data were to be licensed under [CC\ BY\ 4.0](https://creativecommons.org/licenses/by/4.0/), all maps produced from the data would likely need to fulfill this license's obligation for unrestricted access if shared [see @Poole2017]. Similarly, following an example from @Matt2009, if one were to choose the copyleft license [CC\ BY-SA\ 4.0](https://creativecommons.org/licenses/by-sa/4.0/) for this purpose, any map that is a derivative of the data would also need to licensed under CC\ BY-SA\ 4.0 (or a compatible license) if shared. If the intention is to only have [derivative databases](https://osmfoundation.org/wiki/Licence/Community_Guidelines/Produced_Work_-_Guideline) under the same license, one might want to choose the [ODbL\ 1.0](https://opendatacommons.org/licenses/odbl/summary/) for the database, as it was specifically designed not to apply to works produced from the data in the database. Otherwise [CC0\ 1.0](https://creativecommons.org/publicdomain/zero/1.0/) is an excellent choice for data.
- __works in the public domain__: If a work is already in the public domain, it should be marked using the [PDM 1.0](https://creativecommons.org/publicdomain/mark/1.0/), rather than applying a waiver such as the [CC0\ 1.0](https://creativecommons.org/publicdomain/zero/1.0/) (or even another license).
:::

@@

The consequence of applying a permissive license to your work is that others may distribute their adaptations from it under stricter terms, possibly without a free/open license. You may prefer that everybody who receives somebody else's adaptations of your work enjoys the same rights that you granted to them. This can be achieved by _copyleft_ licenses.^[So called because copyleft licenses make content permanently free, thus turning copyright around: © → 🄯. In the realm of Creative Commons licenses, copyleft is also called _share alike_.] For software, they come in two flavors:

@@

The answer is that you need to indicate the license on a per-file or per-folder basis (rather than choosing one for the whole project). Sometimes, even different parts of a file might be subject to different licenses. The easiest approach is to make note of every foreign work included in your project and record its license. Then, multi-license all the remaining files, which are yours, under a code and a non-code license in parallel as explained in @tip-license-r.
